---
title: "My SQL"
author: "Tyler Neumann"
date: "5/1/2022"
output: html_document
---

```{r libs}
library(dbUpsert)
```

```{r rm_con}
if (exists("db_con") == TRUE){
  dbDisconnect(db_con)
  rm(db_con)
}
```

## Start MySQL Docker Container ##

First, stop and remove container if it is currently running.

```{bash stop_container}
docker ps -q --filter "name=eg_mysql" | grep -q . && docker stop eg_mysql && docker rm -fv eg_mysql
```

Then start the container.

```{bash start_container}
docker run --name eg_mysql -p 3306:3306 -e MYSQL_ROOT_PASSWORD=My1secretpassword -e MYSQL_DATABASE=main -d mysql
```

## Open Connection to MySQL ##

The MySQL Docker container take a few seconds to complete the start up process before being able to connect. We must wait ~15 seconds before trying to connect.

```{r sql_con}
Sys.sleep(15)

db_con <- dbConnect(
  RMySQL::MySQL(),
  host = "localhost",
  user = "root",
  password = "My1secretpassword",
  port = 3306,
  dbname = "main"
)

db_info <- list(
  rdbms = db_con |> class() |> as.character(),
  version = db_con |> dbVersion()
)

db_info
```

### MySQL Local Infile ###

MySQL must have this option enabled to write data from R session to the server.

```{sql my_sql_local_infile, connection = db_con}
SET GLOBAL local_infile = 1
```

## Validate DB Interface ##

```{r write_table}
df1 <- data.frame(
  var_one = 1:10,
  var_two = rnorm(10)
)

dbWriteTable(
  conn = db_con,
  name = "test_one",
  value = df1,
  overwrite = T
)
```

```{r read_table}
dbReadTable(db_con, "test_one")
```
